\chapter{Introduction}\label{chapter:introduction}

% TODO: Citations

% TODO(Antoine): You answer a lot of "what" questions.  For an introduction, the "why" questions are maybe more important.  "Why mechanized meta-theory?"  "Why dependent types?"  "Why structural editing?"  As a reader, I don't know what problems are solved with these tools.

% What is mechanized meta-theory?

% What are dependently-typed programming languages?

% What is Beluga? What is HOAS? What is contextual modal type theory?

\Beluga~\cite{pientka2010beluga} is a dependently-typed programming language for specifying and proving properties about formal systems using contextual modal type theory~\cite{nanevski2008contextual}.
It leverages the \acf{LF}~\cite{harper1993framework}, extended with explicit contexts, contextual objects and substitutions~\cite{DBLP:journals/corr/abs-1009-2789, cave2013first}, which allow for logic reasoning involving open terms.
Using \ac{HOAS}~\cite{pfenning1988higher}, elegant and succinct datatype definitions can be made in \LF to subsequently prove theorems pertaining to programming languages and $ \lambda $-calculi.
These proofs are encoded as programs written in a functional style, together with semantic checkers to ensure they always terminate and cover all possible cases.
Propositions then correspond to types, and well-founded reasoning by induction corresponds to recursion.
With this framework, \Beluga has notably been used to mechanically verify proofs for the type safety of \SystemF, cut-admissibility in the intuitionistic sequent calculus, and termination of weak-head normalization for the simply-typed lambda calculus.
Extensions to \Beluga have facilitated the development of proofs by providing interactive tools for manipulating and constructing proof terms.

% How does one specify a formal system using Beluga?

% What are interactive theorem provers?

Interactive theorem proving is the computer-assisted process of finding valid proofs for propositions in a logic system.
Proof assistants have been implemented and successfully leveraged to prove fundamentals of mathematics and type theories.
\Agda~\cite{clffolp, norell2007towards, agda2023}, \Coq~\cite{Coq, bertot2013interactive}, \Isabelle~\cite{nipkow2002isabelle} and \Lean~\cite{lean4} are among the most prevalent proof assistants used in the formalization of programming languages.
Functionalities of typical interactive theorem provers range from \acp{REPL} allowing for structural editing and code inspection to automated proof search to generate valid programs.

% What is Harpoon?

\Harpoon~\cite{errington2021harpoon} is a command-line frontend to \Beluga for proving theorems interactively.
It provides a proof development workflow closer to proofs on paper, with built-in commands replicating proof techniques such as case analyses and appeals to induction hypotheses.
Additional administrative commands are supported to navigate through the history of input commands, and to checkout different holes in proofs declared elsewhere in \Beluga signatures.
Upon closing interactive sessions, \Harpoon produces verbose yet simple tree-shaped proof scripts which can be translated into well-formed \Beluga programs.
\Harpoon's design is largely inspired by \Coq's interactive proof mode using tactics~\cite{delahaye2000tactic} to solve unproven goals.
Crucially, \Harpoon is a system integrated into \Beluga's core functionalities in such a way that the overall proof state does not have to be reconstructed from scratch on every command, which is akin to structural editing.

% What is structured editing?

Structural (or projectional) editing in programming language tooling is the functionality of a software editor that allows the user to manipulate \acp{AST} directly.
This is in contrast with plain text editors in which users may only edit textual representations of their programs, that then subsequently require parsing to be interpreted.
The \Hazel~\cite{omar2017hazelnut, omar2019live} functional programming environment is a notable example of a structural editor in the functional programming community for its ability to dynamically provide type information as the user writes programs with holes.
This is achieved using carefully designed semantics of moving a text-editing cursor in the concrete syntax and mapping its location to a node in the parsed \ac{AST}.
When it comes to theorem provers, one example of structured editing is the \CoqIDE, which leverages the \Coq \XML protocol and a state transaction system to interact with the proof system's kernel by way of messages.
This is not too dissimilar to what is done with implementations of the language server protocol to provide advanced editing functionalities for programming languages.
In essence, the main objectives for implementing structural editing are to enable and assist in the incremental development of programs.
This is in contrast with incremental compilation, which focuses on recompiling minimal sets of programs when they are affected by changes to the source code, which is achieved using dependency analysis and careful handling of cache invalidations.

% What are typical edit actions in programming languages?

Structural editors define sets of edit actions the user may execute at given points in their editing session.
Typical edit actions include navigating the \ac{AST} being modified, or constructing new nodes using either a graphical interface or a textual language.
These actions are then reflected in the editor state in a predictable way, such that the edited program does not need to be reparsed from scratch.
Having a greater control over how a program is edited is also used to prevent or signal edit actions that can invalidate a program.
This can be as simple as signalling type errors on-the-fly, or in the case of theorem provers, to reject unsound operations like invalid appeals to induction hypotheses.

% What are edit actions required by Beluga and Harpoon?

Although \Beluga's interactive mode and \Harpoon are \acp{REPL} as opposed to full-fledged structural editors, they do share some functional requirements.
Indeed, like in \Isabelle, \Hazel and \Coq, users of \Beluga and \Harpoon postpone the completion of programs or proof scripts by inserting holes in them.
These holes stand for missing expressions, and they are later filled in with the help of a proof assistant that provides typing information for the identifiers in scope for each hole.
As such, both \acp{REPL} and structural editors are required to construct and update an editor state as the user performs edit actions.

% In general, what are challenges to implementing edit actions?

Edit actions in \Beluga and \Harpoon include navigating between holes in programs and proof scripts respectively, as well as displaying type information for expressions, and performing case analyses on meta-level and computation-level objects identified by (meta-)variables.
These actions require surgical manipulation of the editor state to ensure soundness of the \ac{AST} being edited is preserved, as well as to prevent the undesirable propagation of information like typing constraints in unrelated edit locations.

% In Beluga and Harpoon specifically, what are the challenges to implementing edit actions?

Historically, \Beluga was not implemented with the mindset of supporting the level of interactivity that \Harpoon purports to have.
Indeed, as is the case with prototype implementations of programming languages, \Beluga's early development focused on implementing a parser for its concrete syntax and a type-checker for its type theory.
The software architectural pattern that arose from this implementation was the pipeline pattern, in which the data processing of programs from its textual to its \ac{AST} representation is handled in single-responsibility phases that sequentially augment and refine the data for later use.
Specifically with respect to \Beluga, these phases have evolved to include implicit argument reconstruction, the translation from a named to a nameless representation of binders using de Bruijn indices, normalization, type-checking, and totality-checking.
What transpires from this design is a unilateral flow of information from the concrete syntax to the internal representation of programs in \Beluga.
Early optimizations were put in place to maximize the performance of this processing pipeline.
In particular, a single global mutable representation of the \Beluga program being processed was implemented to provide information to later processing phases in a feedforward fashion.
This simplified the implementation of new features since globally accessible data does not need intricate data routing procedures like dependency injection.
This design was sound only under the assumption that the necessary information for processing \Beluga signatures would flow in only one direction.
Unfortunately, the introduction of \Beluga's interactive mode and subsequently \Harpoon inadvertently broke that assumption.
Indeed, soundness issues have arose with regards to features and processes of the two systems that require inspecting only a subset of a signature, and as such using referencing environments built out of order.
This issue of locality in references has further been shown to deteriorate the performance of \Beluga's logic programming engine used to automate proof search.

% What is the proposed approach to supporting edit actions in Beluga and Harpoon?

%In particular, we present an explicitly stateful algorithm for the parsing, disambiguation and indexing of \Beluga programs to ensure correct name resolutions across editing states.
%This fixed the soundness issue for \Harpoon proof sessions whereby constants declared after the proof hole would be in scope and shadow earlier declarations.
%The revised approach to name resolution allowed for consistent namespacing of constants to be implemented.
%Additionally, the \Beluga feature of fixity pragmas for defining prefix, infix or postfix notations for \LF-level constants was expanded upon to also work with computation-level constants.
%Using the proposed specification for \Beluga and \Harpoon's grammars, the generation of \HTML pages corresponding to signatures was restored, with a new and improved pretty-printing algorithm of the concrete syntax.
%Overall, the changes have improved the maintainability, resiliency and stability of the affected modules, and have proposed better architectural design patterns that may be applied to the rest of the implementation.

This thesis reports on improvements made to the implementation \Beluga and \Harpoon to address some of the soundness issues in interactive proof development in these two systems.
Overall, the changes listed below have improved the maintainability, resiliency and stability of the affected modules.
Better architectural software design patterns have been put in place, and these may be expanded upon in future iterations on the later phases of semantic analysis.
The specific contributions of this work are:
\begin{enumerate}
\item
A complete and formal specification of \Beluga and \Harpoon's grammars for parsing.
\item
A robust implementation of a context-free parser for \Beluga and \Harpoon, along with a new context-sensitive disambiguation phase to rectify and expand existing features.
Most notably, the \Beluga feature of fixity pragmas for defining prefix, infix or postfix notations for \LF-level constants is reworked to also support computation-level constants.
\item
A uniform name resolution algorithm to improve the clarity of \Beluga programs, implement namespaces, and to support sound incremental proof development.
This fixed soundness issues for \Harpoon having to do with its feature of navigation between holes in proofs anywhere in a \Beluga signature.
\item
An improved and context-aware pretty-printing algorithms for formatting \Beluga programs in external \ac{AST} representation, and for generating corresponding \HTML pages.
\end{enumerate}

% TODO: To justify the addition of modules (functioning only as namespaces), use an example of translation between two languages (deBruijn vs. HOAS for instance)

% TODO: To justify the dynamic side of things, consider a query appearing in the middle of a signature. The logic engine is considering too large of a search space by including the constants appearing later in the signature.

% TODO: To additionally justify the dynamic side of things, talk about structured proof editing in a dynamic environment like Harpoon, or Beluga's interactive mode.
